---
title: "section_C"
author: "Daniel Suh"
date: "10/29/2021"
output: html_document
---
```{r global options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(sf)  
library(terra)  
library(tidyverse)  
library(spData)
library(magrittr)
```


## Spatial Data Operations

Spatial objects can be described using their location and shape and these characteristics are used for most spatial data operations. In the case of vector data, many of these operations are analogous to their attribute data counterpart. Subsetting, joining, and aggregation use similar functions but there are additional decisions to be made when dealing with vector data. For example, if you want to know if a point lies inside a polygon then you will need to decide if a point that intersects with the border of that polygon still counts and if you want to include any points that are close to being inside the polygon. For raster data we can subset cells in similar ways to attribute data and we will also learn about map algebra and the different classes of raster operations: local, focal, zonal, and global.

## Spatial subsetting

Spatial subsetting is the process of selecting features of a spatial object based on whether or not they in some way relate in space to another object. It is analogous to attribute subsetting and can be done with the base R operator [] or with the 'dplyr' filter() function. The 'nz' and 'nz_height' datasets can be used to demonstrate this.

```{r}
canterbury = nz %>% filter(Name == "Canterbury")
canterbury_height = nz_height[canterbury, ]
```






E1. It was established in Section 4.2 that Canterbury was the region of New Zealand containing most of the 100 highest points in the country. How many of these high points does the Canterbury region contain?

```{r}
nz_high_points = nz %>%
  st_join(nz_height) %>%
  group_by(Name) %>%
  summarize(count = n())
```


E2. Which region has the second highest number of nz_height points in, and how many does it have?

West Coast


E3. Generalizing the question to all regions: how many of New Zealand’s 16 regions contain points which belong to the top 100 highest points in the country? Which regions?

```{r}

```


Bonus: create a table listing these regions in order of the number of points and their name.
E4. Use data(dem, package = "spDataLarge"), and reclassify the elevation in three classes: low, medium and high. Secondly, attach the NDVI raster (data(ndvi, package = "spDataLarge")) and compute the mean NDVI and the mean elevation for each altitudinal class.

```{r}
data(dem, package = "spDataLarge")
data(ndvi, package = "spDataLarge")

```


E5. Apply a line detection filter to raster(system.file("external/rlogo.grd", package = "raster")). Plot the result. Hint: Read ?raster::focal().

E6. Calculate the NDVI of a Landsat image. Use the Landsat image provided by the spDataLarge package (system.file("raster/landsat.tif", package = "spDataLarge")).

E7. A StackOverflow post shows how to compute distances to the nearest coastline using raster::distance(). Retrieve a digital elevation model of Spain, and compute a raster which represents distances to the coast across the country (hint: use getData()). Second, use a simple approach to weight the distance raster with elevation (other weighting approaches are possible, include flow direction and steepness); every 100 altitudinal meters should increase the distance to the coast by 10 km. Finally, compute the difference between the raster using the Euclidean distance and the raster weighted by elevation. Note: it may be wise to increase the cell size of the input raster to reduce compute time during this operation.